---
title: "Hands-on Exercise 1"
author: "Serene Cheng"
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()
```
# 1.3.1 R Graphics VS ggplot
## R Graphics
```{r}
hist(exam_data$MATHS)
```

## ggplot2
```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## Essential Grammatical Elements in ggplot2: data
```{r}
ggplot(data=exam_data)
```

A blank canvas appears.

## Essential Grammatical Elements in ggplot2: Aesthetic mappings
This will add the aesthetic mappings.
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

## Geometric Objects: geom_dotplot
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```
The y scale not very useful, in fact misleading. Code below will turn off the y-axis and change the binwidth
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## Geometric Objects: geom_histogram()
```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```