---
title: "Hands-on Exercise 10"
author: "Serene Cheng"
date: "June 25, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

# Reflections

I learned that information dashboards are not just about packing many charts onto one page, but about choosing visual idioms - such as bullet charts and sparklines - that let a viewer judge performance against a target or trend at a glance. Bullet charts were a useful reminder that a single bar can encode several reference bands (poor, satisfactory, good) plus a target marker, all without needing a separate legend. Sparklines showed me how `gt`/`gtExtras` and `reactable`/`reactablefmtr` take the same idea further by embedding a miniature trend line directly inside a table cell, which keeps the table compact while still conveying the shape of the data over time. Comparing the static (`gt`/`gtExtras`) and interactive (`reactable`/`reactablefmtr`) approaches also helped me see the trade-off between a dashboard meant for a printed/exported report versus one meant to be explored on screen.

# Introduction

By the end of this hands-on exercise, we will be able to:

-   Create a bullet chart using **ggplot2**.

-   Create sparklines using **ggplot2**.

-   Build a static information dashboard using **gt** and **gtExtras**.

-   Build an interactive information dashboard using **reactable** and **reactablefmtr**.

# Getting Started

The following R packages will be used for this hands-on exercise.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse)
```

-   **tidyverse** provides a collection of packages for importing, tidying, wrangling and visualising data, including **readr**, **tidyr**, **dplyr**, **ggplot2**, **tibble**, **stringr**, **forcats** and **purrr**.

-   **lubridate** provides functions to work with dates and times more efficiently.

-   **ggthemes** is an extension of **ggplot2** that provides additional themes beyond the basic ones.

-   **gtExtras** provides helper functions to assist in creating tables with **gt**, a package for making polished tables in R.

-   **reactable** provides functions to create interactive data tables for R, based on the React Table library.

-   **reactablefmtr** streamlines and enhances the styling of interactive **reactable** tables with customisable functions and themes.

# Importing Microsoft Access Database

## The data set

For this exercise, a personal database in Microsoft Access mdb format called *Coffee Chain* is used.

## Importing database into R

`odbcConnectAccess()` of **RODBC** package is used to import a database query table into R.

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

**Note:** before running the code chunk above, the R system needs to be switched to the 32-bit version, because `odbcConnectAccess()` is based on 32-bit, not 64-bit. As the rds file has already been produced, this chunk is not re-run here.

## Data Preparation

The code chunk below imports `CoffeeChain.rds` into R.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

The code chunk below aggregates Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## Bullet chart in ggplot2

The code chunk below plots a bullet chart using **ggplot2** functions.

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current),
           width = 0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax = target),
                width = .4,
                colour = "red",
                linewidth = 1) +
  coord_flip()
```

# Plotting Sparklines Using ggplot2

## Preparing the data

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

The code chunk below computes the minimum, maximum and end-of-month sales.

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))
```

The code chunk below computes the 25th and 75th quantiles.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)
```

## Sparklines in ggplot2

```{r}
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, ymax = quart2),
              fill = 'grey90') +
  geom_line(linewidth = 0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 7, base_family = "Helvetica") +
  theme(axis.title = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

# Static Information Dashboard Design: gt and gtExtras Methods

In this section, a static information dashboard is created using **gt** and **gtExtras** packages.

## Plotting a simple bullet chart

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
                target = target,
                width = 60,
                palette = c("lightblue",
                            "black")) %>%
  gt_theme_538()
```

# Sparklines: gtExtras Method

Before preparing the sales report by product using **gtExtras** functions, the code chunk below prepares the data.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date,
                        label = TRUE,
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

`gtExtras` functions almost exclusively require a data frame with list columns. The code chunk below converts the `report` data frame into list columns.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

## Plotting Coffeechain Sales report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Adding statistics

```{r}
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = TRUE),
            "Max" = max(Sales, na.rm = TRUE),
            "Average" = mean(Sales, na.rm = TRUE)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
             decimals = 2)
```

## Combining the data.frame

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")

sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = TRUE),
            "Max" = max(Sales, na.rm = TRUE),
            "Average" = mean(Sales, na.rm = TRUE)
            )

sales_data <- left_join(sales, spark)
```

## Plotting the updated data.table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Combining bullet chart and sparklines

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()

sales_data <- sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue",
                            "black")) %>%
  gt_theme_538()
```

# Interactive Information Dashboard Design: reactable and reactablefmtr Methods

In this section, an interactive information dashboard is created using **reactable** and **reactablefmtr** packages.

To build interactive sparklines, the **dataui** R package is required.

```{r}
#| eval: false
remotes::install_github("timelyportfolio/dataui")
```

```{r}
library(dataui)
```

## Plotting interactive sparklines

Similar to **gtExtras**, plotting interactive sparklines with **reactablefmtr** requires preparing a list field first.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

`react_sparkline()` is used to plot the sparklines.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Changing the pagesize

By default the page size is 10. The `defaultPageSize` argument changes this setting.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Adding points and labels

`highlight_points` shows the minimum and maximum value points, and `labels` labels the first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## Adding reference line

The `statline` argument shows the mean line.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## Adding bandline

Instead of a reference line, a bandline can be added using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## Changing from sparkline to sparkbar

Instead of displaying the values as sparklines, they can be displayed as sparkbars.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
