---
title: "Hands-on Exercise 5b"
author: "Serene Cheng"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Reflections

I learned that correlation matrices help us quickly understand the relationship and strength between many numerical variables at the same time. I also learned how to use pairs() in R to create scatterplot matrices, but it can become cluttered when there are many observations. Corrgram tools such as ggcorrmat() and corrplot() make the relationships easier to interpret using colours, shapes, numbers, and significance tests. Overall, this exercise helped me see how correlation visualisation can support exploratory data analysis before building models or deeper analysis.

# Introduction

A correlation coefficient measures the direction and strength of the relationship between two variables, ranging from -1 to 1.

When there are many variables, their pairwise relationships can be shown in a correlation matrix or scatterplot matrix. This helps us understand patterns, use the results in further analysis, and check for issues such as highly correlated variables in regression.

For large datasets, a corrgram is useful because it visualises correlations using colours, shapes, or ordering, making patterns easier to see.

In this hands-on exercise, we will learn how to create correlation matrices in R using pairs(), visualise corrgrams using corrplot, and build an interactive correlation matrix using plotly.

# Getting started

## Installing and loading R packages

Install and launch **corrplot**, **ggpubr**, **plotly** and **tidyverse** in RStudio.

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

## Data importing

[Wine Quality Data Set](https://archive.ics.uci.edu/ml/datasets/wine+quality) of UCI Machine Learning Repository will be used. The data set consists of 13 variables and 6497 observations. For the purpose of this exercise, we have combined the red wine and white wine data into one data file. It is called wine_quality and is in csv file format.

read_csv() function will be used for importing. Beside the quality and type, the rest of the variables are numerical and continuous data type.

```{r}
wine <- read_csv("data/wine_quality.csv")
```

# Building Correlation Matrix: pairs() method

There's more than 1 way to build scatterplot matrix with R. We will learn how to create scatterplot matrix in this section using the pairs function of R graphics.

## Building a basic correlation matrix

Figure below shows the scatter plot matrix of Wine Quality Data. It is a 11 by 11 matrix.

```{r}
pairs(wine[,1:11])
```

Required input of *pairs()* can be a matrix or data frame. The code chunk used to create the scatterplot matrix is quite simple. It uses the default *pairs* function. Columns 2 to 12 of wine dataframe is used to build the scatterplot matrix. The variables are: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates and alcohol.

```{r}
pairs(wine[,2:12])
```

## Drawing the lower corner

pairs function in R also provides many customisation arguments. It is a common practise to either show the upper or lower half of the correlation matrix instead of both. This is because a correlation matrix is symmetric.

To show the lower half of the correlation matrix, the upper.panel argument will be used as shown in the code chunk below.

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

Displaying upper half:

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

## Including with correlation coefficients

[*panel.cor*](https://www.rdocumentation.org/packages/xcms/versions/1.48.0/topics/panel.cor) function will be used. This will also show higher correlations in a larger font.

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

# Visualising Correlation Matrix: ggcorrmat() method

One of the major limitation of the correlation matrix is: the scatter plots appear very cluttered when the number of observations is relatively large (i.e. more than 500 observations). To overcome this problem, **Corrgram** data visualisation technique suggested by D. J. Murdoch and E. D. Chow (1996) and Friendly, M (2002) and will be used.

There are at least three R packages provide function to plot corrgram:

-   [corrgram](https://cran.r-project.org/web/packages/corrgram/index.html)

-   [ellipse](https://cran.r-project.org/web/packages/ellipse/index.html)

-   [corrplot](https://cran.r-project.org/web/packages/corrplot/index.html)

We will learn how to visualising correlation matrix by using [*ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggcorrmat.html) of [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/index.html) package.

## Basic plot

One of the advantage of using *ggcorrmat()* over many other methods to visualise a correlation matrix is it’s ability to provide a comprehensive and yet professional statistical report as shown in the figure below.

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)
```

-   `cor.vars` argument is used to compute the correlation matrix needed to build the corrgram.

-   `ggcorrplot.args` argument provide additional (mostly aesthetic) arguments that will be passed to [`ggcorrplot::ggcorrplot`](http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2) function. The list should avoid any of the following arguments since they are already internally being used: `corr`, `method`, `p.mat`, `sig.level`, `ggtheme`, `colors`, `lab`, `pch`, `legend.title`, `digits`.

The sample sub-code chunk can be used to control specific component of the plot such as the font size of the x-axis, y-axis, and the statistical report.

```{r}
ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

# Building multiple plots

`ggcorrmat()` = one overall correlation chart `grouped_ggcorrmat()` = separate correlation charts for each group/category, like red wine vs white wine.

```{r}
#| fig-width: 12
#| fig-height: 14
#| fig-align: center

grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 1),
  ggcorrplot.args = list(
    outline.color = "black", 
    hc.order = TRUE,
    tl.cex = 12
  ),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

-   to build a facet plot, the only argument needed is `grouping.var`.

-   Behind *group_ggcorrmat()*, **patchwork** package is used to create the multiplot. `plotgrid.args` argument provides a list of additional arguments passed to [*patchwork::wrap_plots*](https://patchwork.data-imaginist.com/reference/wrap_plots.html), except for guides argument which is already separately specified earlier.

-   Likewise, `annotation.args` argument is calling [*plot annotation arguments*](https://patchwork.data-imaginist.com/reference/plot_annotation.html) of patchwork package.

# Visualising Correlation Matrix using corrplot Package

We will focus on corrplot. [An Introduction to corrplot Package](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) can be explored to gain basic understanding of **corrplot** package. The other two packages are corrgram and ellipse

## Getting started with corrplot

Before we can plot a corrgram using *corrplot()*, we need to compute the correlation matrix of wine data frame.

[*cor()*](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/cor) of R Stats is used to compute the correlation matrix of wine data frame below

```{r}
wine.cor <- cor(wine[, 1:11])
```

Next, [*corrplot()*](https://www.rdocumentation.org/packages/corrplot/versions/0.2-0/topics/corrplot) is used to plot the corrgram by using all the default setting as shown in the code chunk below.

```{r}
corrplot(wine.cor)
```

By default, `corrplot()` uses **circles** to show correlations in a **symmetric matrix**. **Blue** represents positive correlations, while **red** represents negative correlations. The **darker the colour**, the stronger the relationship; the **lighter the colour**, the weaker the relationship.

## Working with visual geometrics

There are 7 visual geometrics (parameter method) that can be used to encode attribute values. They are: circle, square, ellipse, number, shade, color and pie. The default is circle. As shown in the previous section, the default visual geometric of corrplot matrix is circle. However, this default setting can be changed by using the *method* argument as shown in the code chunk below

```{r}
corrplot(wine.cor, 
         method = "ellipse") 
```

Now, lets try pie.

```{r}
corrplot(wine.cor, 
         method = "pie") 
```

## Working with layout

*corrplot()* supports three layout types, namely: “full”, “upper” or “lower”. The default is “full” displaying full matrix. The default setting can be changed by using the *type* argument of *corrplot()*.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower")
```

The default layout of the corrgram can be further customised. For example, arguments *diag* and *tl.col* are used to turn off the diagonal cells and to change the axis text label colour to black colour respectively as shown in the code chunk and figure below.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower",
         diag = FALSE,
         tl.col = "black")
```

Now lets try other methods - tl.pos

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.pos = "lt",
         tl.cex = 0.8,
         tl.offset = 0.8,
         cl.pos = "r",
         cl.cex = 0.8,
         cl.offset = 0.5)
```

## Working with mixed layout

With **corrplot** package, it is possible to design corrgram with mixed visual matrix of one half and numerical matrix on the other half. In order to create a corrgram with mixed layout, the [*corrplot.mixed()*](https://www.rdocumentation.org/packages/corrplot/versions/0.84/topics/corrplot.mixed), a wrapped function for mixed visualisation style will be used.

Figure below shows a mixed layout corrgram plotted using wine quality data.

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The `lower` and `upper` arguments control how each half of the corrgram is displayed. In this example, the **lower half** uses ellipses, while the **upper half** shows correlation values as numbers. The `tl.pos` argument controls where the axis labels appear, and `diag` controls what is shown on the diagonal line of the corrgram.

## Combining corrgram with the significant test

In statistical analysis, we are also interested to know which pair of variables their correlation coefficients are statistically significant.

Figure below shows a corrgram combined with the significant test. The corrgram reveals that not all correlation pairs are statistically significant. For example the correlation between total sulfur dioxide and free surfur dioxide is statistically significant at significant level of 0.1 but not the pair between total sulfur dioxide and citric acid.

With corrplot package, we can use the *cor.mtest()* to compute the p-values and confidence interval for each pair of variables.

```{r}
wine.sig = cor.mtest(wine.cor, conf.level= .95)
```

```{r}
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

## Reorder a corrgram

Reordering a correlation matrix helps reveal hidden patterns by placing similar variables closer together. By default, `corrplot()` keeps the original variable order, but we can change this using the `order` argument. The package supports sorting methods such as

-   “AOE” is the angular order of the eigenvectors. See Michael Friendly (2002) for details.

-   “FPC” for the first principal component order.

-   “hclust” for hierarchical clustering order, and “hclust.method” for the agglomeration method to be used.

    -   “hclust.method” should be one of “ward”, “single”, “complete”, “average”, “mcquitty”, “median” or “centroid”.

-   “alphabet” for alphabetical order.

More algorithms can be found in **seriation** package.

```{r}
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```

## Reordering a correlation matrix using hclust

***corrplot()*** can draw rectangles around the corrgram based on the results of hierarchical clustering.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```

# References

Michael Friendly (2002). “Corrgrams: Exploratory displays for correlation matrices”. The American Statistician, 56, 316–324.

D.J. Murdoch, E.D. Chow (1996). “A graphical display of large correlation matrices”. The American Statistician, 50, 178–180.
