---
title: "Hands-on Exercise 5d"
author: "Serene Cheng"
date: "May 14, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Reflections

I learnt that parallel coordinates plots are useful for visualising and comparing many numerical variables at the same time. I also learnt how static plots can be created using ggparcoord() and improved with grouping, scaling, boxplots, faceting, and rotated labels. It also taught me how interactive plots using parallelPlot() can make multivariate patterns easier to explore.

# Introduction

A **parallel coordinates plot** is used to visualise and analyse data with many numerical variables at the same time. It helps compare multiple variables and identify relationships or patterns between them, such as factors contributing to the Happiness Index. However, it is more useful for **interactive analysis and exploring patterns** rather than for simple business presentations or general audiences.

We will learn below by the end of this:

-   plotting statistic parallel coordinates plots by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://r4va.netlify.app/chap15) package, and

-   plotting interactive parallel coordinates plots by using [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) package.

# Getting started

## Installing and loading R packages

**GGally**, **parallelPlot** and **tidyverse** packages will be used.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## Data importing

[World Happines 2018 report](https://worldhappiness.report/ed/2018/) will be used. The data set is downloaded from [here](https://s3.amazonaws.com/happiness-report/2018/WHR2018Chapter2OnlineData.xls). The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

# Plotting Static Parallel Coordinates Plot

We will learn how to plot static parallel coordinates plot by using ggparcoord() of GGally package. Before getting started, it is a good practice to read the function description in detail

## Plotting a simple parallel coordinates

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using ggparcoord().

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

Only two argument namely `data` and `columns` is used. `Data` argument is used to map the data object (i.e. *wh*) and `columns` is used to select the columns for preparing the parallel coordinates plot.

## Plotting a parallel coordinates with boxplot

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. In this section, we will learn how to makeover the plot by using a collection of arguments provided by `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

Learning points from above:

-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

-   `alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

-   `title` argument is used to provide the parallel coordinates plot a title.

## Parallel coordinates with facet

Since `ggparcoord()` is developed by extending **ggplot2** package, we can combine the use of some of the ggplot2 function when plotting a parallel coordinates plot.

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia. But some of the variables names overlap on x-axis

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

## Rotating x-axis text label

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

-   To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. And we specify `element_text(angle = 30)` to rotate the x-axis text by an angle 30 degree.

## Adjusting rotated x-axis text label

We can avoid label overlap with plot by adjusting the text location using `hjust` argument to theme’s text element with `element_text()`. We use `axis.text.x` as we want to change the look of x-axis text

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

# Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and [d3.js](https://d3js.org/). In this section, we will learn how to use functions provided in parallelPlot package to build interactive parallel coordinates plot.

## Basic plot

Plot interactive parallel coordinates using `parallelPlot()`.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

Some of the axis labels are too long. 

## Rotate axis label

In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

-   One useful interactive feature of `parallelPlot` is that we can click on a variable of interest, such as Happiness Score. The default monotonous blue colour will then change to a blue colour scheme with different intensity levels.

## Changing colour scheme

Change the default blue colour scheme by using `continousCS` argument as shown in the code chunl below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## Parallel coordinates plot with histogram

In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

# References

-   [*ggparcoord()*](http://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](http://ggobi.github.io/ggally/index.html) package

-   [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)
