---
title: "Hands-on Exercise 9a"
author: "Serene Cheng"
date: "June 19, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

# Reflections

I learned how to create choropleth maps in R using the **tmap** package, which is purpose-built for thematic mapping and offers far more cartographic control than general-purpose plotting packages. The quickest way to get a map on screen is `qtm()`, but building maps layer by layer with `tm_shape()`, `tm_polygons()`, `tm_fill()` and `tm_borders()` gives much finer control over classification, colour and layout. I also learned that the choice of data classification method (e.g. quantile vs equal) can change the visual story a map tells even when the underlying data is identical, which is an important reminder that "maps lie". Finally, I practiced preparing small multiple choropleth maps in three different ways and learned how to subset spatial objects to map only a selected criterion.

# Introduction

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colours. By the end of this hands-on exercise, we will be able to:

-   Import geospatial data in ESRI shapefile format using `st_read()` of **sf** package.

-   Import aspatial data in csv file format using `read_csv()` of **readr** package.

-   Perform data wrangling using functions of **dplyr** package.

-   Plot quick and highly customisable choropleth maps using `qtm()` and **tmap**'s elements.

# Getting Started

In this hands-on exercise, the key R package used is **tmap**. Beside **tmap**, four other R packages will be used:

-   **readr** for importing delimited text files,

-   **tidyr** for tidying data,

-   **dplyr** for wrangling data, and

-   **sf** for handling geospatial data.

Among these, **readr**, **tidyr** and **dplyr** are part of **tidyverse**, so only **tidyverse** needs to be installed rather than each one individually.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Importing Data into R

## The Data

Two data sets will be used to create the choropleth map:

-   Master Plan 2014 Subzone Boundary (Web) (i.e. `MP14_SUBZONE_WEB_PL`) in ESRI shapefile format, downloaded from [data.gov.sg](https://data.gov.sg/). This is a geospatial data set consisting of the geographical boundary of Singapore at the planning subzone level, based on URA Master Plan 2014.

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. `respopagesextod2011to2020.csv`), downloaded from the [Department of Statistics, Singapore](https://www.singstat.gov.sg/). This is an aspatial data file - although it does not contain coordinate values, its PA and SZ fields can be used as unique identifiers to geocode against the `MP14_SUBZONE_WEB_PL` shapefile.

## Importing Geospatial Data into R

The code chunk below uses `st_read()` of **sf** package to import the `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/MP14_SUBZONE_WEB_PL",
                layer = "MP14_SUBZONE_WEB_PL")
```

We can examine the content of `mpsz` using the code chunk below.

```{r}
mpsz
```

Notice that only the first ten features are displayed by default.

## Importing Attribute Data into R

Next, we import `respopagesextod2011to2020.csv` into RStudio and save the file into an R data frame called `popdata`, using `read_csv()` of **readr** package.

```{r}
popdata <- read_csv("data/respopagesextod2011to2020.csv")
```

## Data Preparation

Before a thematic map can be prepared, we need a data table with year 2020 values, including the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL and DEPENDENCY:

-   YOUNG: age group 0 to 4 until age group 20 to 24,

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,

-   AGED: age group 65 and above,

-   TOTAL: all age groups, and

-   DEPENDENCY: the ratio between young and aged against the economically active group.

### Data wrangling

The following data wrangling and transformation functions are used: `pivot_wider()` of **tidyr**, and `mutate()`, `filter()`, `group_by()` and `select()` of **dplyr**.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) +
           rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
         / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
         `ECONOMY ACTIVE`, `AGED`,
         `TOTAL`, `DEPENDENCY`)
```

### Joining the attribute data and geospatial data

Before performing the georelational join, the values in the PA and SZ fields need to be converted to uppercase, since the *SUBZONE_N* and *PLN_AREA_N* fields in `mpsz` are already in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of **dplyr** is used to join the geographical data and the attribute table using the planning subzone name (i.e. *SUBZONE_N* and *SZ*) as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                           by = c("SUBZONE_N" = "SZ"))
```

Using `mpsz` as the left data table in `left_join()` ensures that the output remains a simple features data frame.

```{r}
write_rds(mpsz_pop2020, "data/mpszpop2020.rds")
```

# Choropleth Mapping Geospatial Data Using *tmap*

Two approaches can be used to prepare a thematic map using **tmap**: plotting a thematic map quickly with `qtm()`, or plotting a highly customisable thematic map using **tmap**'s elements.

## Plotting a choropleth map quickly by using *qtm()*

`qtm()` is the easiest and quickest way to draw a choropleth map using **tmap**. It is concise and provides a good default visualisation in many cases.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

-   `tmap_mode()` with the "plot" option produces a static map; for interactive mode, "view" should be used instead.

-   The `fill` argument maps the attribute of interest (i.e. DEPENDENCY).

## Creating a choropleth map by using *tmap*'s elements

While `qtm()` is quick, it makes it harder to control the aesthetics of individual layers. To draw a higher quality cartographic choropleth map, **tmap**'s drawing elements are used instead.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

### Drawing a base map

The basic building block of **tmap** is `tm_shape()` followed by one or more layer elements such as `tm_fill()` and `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### Drawing a choropleth map using *tm_polygons()*

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, the target variable (i.e. DEPENDENCY) is simply assigned to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

-   The default interval binning method used is "pretty".

-   The default colour scheme is `YlOrRd` of ColorBrewer.

-   Missing values are shaded grey by default.

### Drawing a choropleth map using *tm_fill()* and *tm_borders()*

`tm_polygons()` is actually a wrapper of `tm_fill()` and `tm_borders()`. `tm_fill()` shades the polygons using the default colour scheme, and `tm_borders()` adds the shapefile's borders onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY")
```

To add the boundary of the planning subzones, `tm_borders()` is used:

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```

`tm_borders()` supports several arguments: `col` for border colour, `lwd` for line width (default 1), and `lty` for line type (default "solid").

## Data classification methods of **tmap**

**tmap** provides ten data classification methods: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher* and *jenks*. The classification method is defined using the *style* argument of `tm_fill()` or `tm_polygons()`.

### Plotting choropleth maps with built-in classification methods

The code chunk below uses the *jenks* classification with 5 classes.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The code chunk below uses the *equal* classification method instead.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The quantile classification distributes the data more evenly across classes than the equal classification method, which is a useful reminder that maps can present the same underlying data very differently depending on the classification chosen.

### Plotting choropleth map with custom breaks

For built-in styles, category breaks are computed internally. To override these defaults, breakpoints can be set explicitly using the *breaks* argument of `tm_fill()`. In **tmap**, breaks include a minimum and maximum, so n+1 elements must be specified to produce n categories.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Based on the summary statistics, break points are set at 0.60, 0.70, 0.80 and 0.90, together with a minimum of 0 and a maximum of 1.00.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

## Colour Scheme

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the **RColorBrewer** package.

### Using ColourBrewer palette

To change the colour, the preferred colour is assigned to the *values* argument of `tm_scale_intervals()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, a "-" prefix is added.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Map Layouts

Map layout refers to the combination of all map elements into a cohesive map, including the objects mapped, title, scale bar, compass, margins and aspect ratio.

### Map Legend

`tm_legend()` provides options to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

### Map style

`tmap_style()` allows a wide variety of layout settings to be changed. The code chunk below uses the *classic* style.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture

`tm_compass()`, `tm_scale_bar()` and `tm_grid()` are used to add a compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

To reset the default style:

```{r}
tmap_style("white")
```

## Drawing Small Multiple Choropleth Maps

Small multiple maps, also known as facet maps, are composed of many maps arranged side-by-side or stacked vertically, enabling visualisation of how spatial relationships change with respect to another variable, such as time. In **tmap**, small multiple maps can be plotted in three ways: by assigning multiple values to at least one aesthetic argument, by defining a group-by variable in `tm_facets()`, or by creating multiple stand-alone maps with `tmap_arrange()`.

### By assigning multiple values to at least one aesthetic argument

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(c("DEPENDENCY", "AGED"),
          style = c("equal", "quantile"),
          palette = list("Blues", "Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### By defining a group-by variable in *tm_facets()*

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### By creating multiple stand-alone maps with *tmap_arrange()*

```{r}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

## Mapping Spatial Objects Meeting a Selection Criterion

Instead of creating small multiple choropleth maps, a selection function can be used to map only spatial objects meeting a selection criterion - in this case, subzones within the Central Region.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

# Reference

## All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)

-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)

-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

## Geospatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)

-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

## Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)

-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
