---
title: "Hands-on Exercise 9b"
author: "Serene Cheng"
date: "June 19, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

# Reflections

I learned how to visualise geospatial point data by converting an aspatial data frame (one with x and y coordinate columns but no spatial geometry) into a simple feature point object using `st_as_sf()` of **sf** package. I also learned how proportional symbol maps differ from choropleth maps - instead of shading enumeration units, the size of a point symbol is scaled to represent the magnitude of a variable, which works well for point-based phenomena such as outlet locations. Lastly, I learned that **tmap**'s interactive "view" mode supports faceted maps with synchronised zoom and pan via the `sync` argument of `tm_facets()`, which makes comparing subgroups side by side much easier.

# Introduction

Proportional symbol maps (also known as graduated symbol maps) use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, such as counts of people. Like choropleth maps, both classed (range-graded) and unclassed (proportional) versions can be created - in the unclassed version, the area of each symbol is proportional to the value of the attribute being mapped. In this hands-on exercise, we will create a proportional symbol map showing the number of wins by Singapore Pools' outlets using the **tmap** package.

By the end of this hands-on exercise, we will be able to:

-   Import an aspatial data file into R.

-   Convert it into a simple point feature data frame, while assigning an appropriate projection reference to the newly created data frame.

-   Plot interactive proportional symbol maps.

# Getting Started

Before getting started, we need to ensure that **tmap** and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Geospatial Data Wrangling

## The Data

The data set used for this hands-on exercise is called `SGPools_svy21`, in csv file format. It consists of seven columns, where the *XCOORD* and *YCOORD* columns are the x-coordinates and y-coordinates of Singapore Pools outlets and branches, expressed in the [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

## Data Import and Preparation

The code chunk below uses `read_csv()` of **readr** package to import `SGPools_svy21.csv` into R as a tibble data frame called `sgpools`.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
```

After importing the data file into R, it is important to examine whether the data has been imported correctly. The code chunk below uses `list()` to do this.

```{r}
list(sgpools)
```

Notice that `sgpools` is a tibble data frame rather than the common R data frame.

## Creating a **sf** data frame from an aspatial data frame

The code chunk below converts the `sgpools` data frame into a simple feature data frame using `st_as_sf()` of **sf** package.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                        coords = c("XCOORD", "YCOORD"),
                        crs = 3414)
```

-   The *coords* argument requires the column name of the x-coordinates first, followed by the column name of the y-coordinates.

-   The *crs* argument requires the coordinate system in EPSG format. [EPSG: 3414](https://epsg.io/3414) is the Singapore SVY21 Projected Coordinate System. Other countries' EPSG codes can be looked up at [epsg.io](https://epsg.io/).

We can display the basic information of the newly created `sgpools_sf` using the code chunk below.

```{r}
list(sgpools_sf)
```

The output shows that `sgpools_sf` is in point feature class with EPSG ID 3414. The bounding box (bbox) provides information on the extent of the geospatial data.

# Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the view mode of **tmap** is used.

The code chunk below turns on **tmap**'s interactive mode.

```{r}
tmap_mode("view")
```

## It all started with an interactive point symbol map

The code chunk below is used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = 1,
             col = "black",
             lwd = 1)
```

## Let's make it proportional

To draw a proportional symbol map, a numerical variable needs to be assigned to the size visual attribute. The code chunk below assigns the variable *Gp1Gp2 Winnings* to the size visual attribute.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Let's give it a different colour

The proportional symbol map can be further improved by also using the colour visual attribute. In the code chunk below, the *OUTLET TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## I have twin brothers :)

An impressive and lesser-known feature of **tmap**'s view mode is that it also works with faceted plots. The *sync* argument of `tm_facets()` can be used to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by = "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Before ending the session, it is good practice to switch **tmap**'s viewer back to plot mode using the code chunk below.

```{r}
tmap_mode("plot")
```

# Reference

## All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)

-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)

-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

## Geospatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)

-   [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

## Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)

-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
